---
title: "Predictive gene selection in asthma"
author: "Damir Zhakparov"
output:
    BiocStyle::html_document:
        toc: true
vignette: >
    %\VignetteIndexEntry{Predictive gene selection in asthma}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

GeneSelectR 2.0 combines repeated gene inclusion and excluded-sample predictive
contribution into a ranking. Both components are compared with shuffled-outcome
references. The output retains individual component measurements for inspection.
The `SummarizedExperiment` input supports use with Bioconductor expression
analysis tools. External biological information is interpreted after selection.

The example uses a small, deterministic subset of GSE69683, a U-BIOPRED blood
expression study. It illustrates the software interface. It does not reproduce
the full asthma benchmark or provide a new estimate of clinical performance.

# Installation

The package is being prepared for submission. Install a local source archive
before running this vignette. After Bioconductor acceptance, installation will
use:

```{r installation, eval=FALSE}
BiocManager::install("GeneSelectR2")
```

# Expression and phenotype

```{r input}
library(GeneSelectR2)
library(SummarizedExperiment)
data(asthma_example)
dim(asthma_example)
table(asthma_example$severity)
metadata(asthma_example)
```

Genes are rows and samples are columns. The phenotype is stored in `colData`.
Expression has already been normalized on a log scale. Real analyses must keep
all preprocessing, variable-gene selection and tuning inside the training data.
An independent test partition is needed to estimate prediction performance.

# Fit the workflow

The small example uses five repeated fits, two outcome permutations and five
fits per permutation, with one worker. These reduced counts keep package checks
short. The published configuration used 50 fits, 20 permutations and 20 null
fits, comparing alpha 0.5 and 1 using internal prediction AUC.

```{r fit}
fit <- select_genes(
    asthma_example, outcome = "severity", n_genes = 20,
    alpha = 0.5, B = 5, permutations = 2, null_B = 5,
    workers = 1, seed = 123
)
fit$alpha_comparison
fit$selected_genes
```

The standard settings can be used with a training experiment as follows:

```{r full-settings, eval=FALSE}
fit <- select_genes(training_experiment, "severity", n_genes = 20)
```

# Inspect gene-level outputs

```{r output}
head(fit$gene_scores)
fit$parameters
```

`recurrence` is inclusion frequency among repeated models.
`predictive_contribution` combines excluded-sample SHAP contribution frequency
and mutual information. `adjusted_recurrence` and `adjusted_contribution`
compare these measurements with shuffled-outcome references. The ratios are
ranking quantities and provide no probability or significance-test
interpretation. The `selected_genes` vector contains the requested top 20
genes.

```{r ranking-plot, fig.width=9, fig.height=5}
plot_gene_ranking(fit, n = 12)
```

# Biological interpretation

Biological information is assessed after predictive selection. It does not
change the gene ranking. Gene Ontology measures functional similarity, Hallmark
sets identify shared biological processes, Open Targets records disease
associations, and STRING supplies protein-association context. Database coverage
and retrieval dates must accompany interpretation. Missing annotation is kept
distinct from a measured score of zero.

The package includes the verified biological summaries used to demonstrate the
workflow. Ratios compare each selected gene set with same-size random gene sets
drawn from the same candidate genes. A value of one denotes the random-set
reference. The measurements remain separate because they describe different
properties.

```{r biology-heatmap, fig.width=8.5, fig.height=6.3}
data(benchmark_biology)
plot_biology_comparison(
    benchmark_biology, method = "GeneSelectR 2.0",
    dataset_order = c(
        "Atopic dermatitis", "Bladder cancer", "Asthma", "Sepsis",
        "Psoriasis", "Tuberculosis", "Crohn disease"
    )
)
```

For asthma, the selected-to-random ratios were 1.65 for GO similarity, 3.75 for
Hallmark sharing, 1.83 for Open Targets at a score cutoff of 0.05, and 1.56 at a
cutoff of 0.10. These are descriptive ratios. They are not probabilities,
p-values or components of the predictive ranking.

# Full-cohort asthma output

The executable fit above uses 96 samples and 300 genes so that the vignette can
be built during package checks. The following figure uses the verified
full-cohort GSE69683 results reported in the manuscript: 323 samples, 15 outer
validation divisions and sets of 20 genes. This packaged data object lets users
inspect and reproduce the display without downloading external databases.

```{r case-study, fig.width=11, fig.height=5.5}
data(asthma_case_study)
plot_gene_evidence(asthma_case_study, comparison_label = "DGE")
```

Outer selection count records membership in each final top-20 gene set.
Within-training recurrence records inclusion among the elastic-net fits used to
construct a ranking. Adjusted contribution is compared with shuffled-outcome
fits. The Open Targets value is a continuous recorded asthma association from
release 26.06. `LINC00537` has no available Open Targets score and remains
marked `NA`.

# Reproducibility

The example source and deterministic subset rule are documented in
`?asthma_example`. Small-sample example results should not be substituted for
the full-cohort results in the manuscript. Internal AUC is used for tuning and
is not an outer-test performance estimate.

```{r session}
sessionInfo()
```
