---
title: "Predictive gene selection in asthma"
author: "Damir Zhakparov"
date: "15 September 2026"
output:
    BiocStyle::html_document:
        toc: true
vignette: >
    %\VignetteIndexEntry{Predictive gene selection in asthma}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

GeneSelectR 2.0 combines repeated gene inclusion and excluded-sample predictive
contribution in a gene ranking. Both components are compared with
shuffled-outcome references. The output retains the individual measurements.
The `SummarizedExperiment` interface supports use after Bioconductor expression
preprocessing. External biological information is evaluated after selection.

The example uses a deterministic subset of GSE69683, a U-BIOPRED blood
expression study. It demonstrates package operation with reduced computation.
The packaged full-cohort result summaries are displayed in later sections.

# Installation

The package is being prepared for Bioconductor submission. After acceptance,
installation will use:

```{r installation, eval=FALSE}
BiocManager::install("GeneSelectR2")
```

# Study design and preprocessing

The outcome in this example compares moderate and severe asthma. The same
interface can represent disease versus control, treatment response, infection
status, or a pre-specified pairwise subtype comparison. The outcome must contain
two factor levels, with the second level treated as positive.

Expression normalization, quality control, batch adjustment, and candidate-gene
selection precede model fitting. Parameters estimated from expression data must
be fitted within each training partition. Independent samples are required for
a predictive-performance estimate. The internal AUC reported by GeneSelectR 2.0
is a tuning measurement from the training workflow.

# Expression and phenotype

```{r input}
library(GeneSelectR2)
library(SummarizedExperiment)
data(asthma_example)
dim(asthma_example)
assayNames(asthma_example)
table(asthma_example$severity)
metadata(asthma_example)
```

Genes are rows and samples are columns. The normalized expression measurements
are stored in the `log_expression` assay. The phenotype is stored in `colData`.
Sample identifiers are retained as column names. Source, platform, expression
scale, and the deterministic subset rule are recorded in `metadata` and in
`?asthma_example`.

```{r inspect-input}
assay(asthma_example, "log_expression")[1:4, 1:4]
head(colData(asthma_example))
head(rownames(asthma_example))
```

# Fit the workflow

The example uses five repeated fits, two outcome permutations, and five fits per
permutation, with one worker. These reduced counts limit package-check time. The
benchmark configuration used 50 fits, 20 permutations, and 20 null fits, with
alpha values 0.5 and 1 compared by internal AUC.

```{r fit}
fit <- select_genes(
    asthma_example,
    outcome = "severity",
    assay_name = "log_expression",
    n_genes = 20,
    alpha = 0.5,
    B = 5,
    permutations = 2,
    null_B = 5,
    workers = 1,
    seed = 123
)
fit$alpha_comparison
fit$selected_genes
```

The standard settings can be applied to a training experiment with:

```{r full-settings, eval=FALSE}
fit <- select_genes(
    training_experiment,
    outcome = "severity",
    assay_name = "log_expression",
    n_genes = 20,
    alpha = c(0.5, 1),
    B = 50,
    permutations = 20,
    null_B = 20,
    workers = 1,
    seed = 123
)
```

# Inspect gene-level outputs

```{r output}
head(fit$gene_scores)
fit$parameters
fit$cv_results[c("mean_auc", "sd_auc", "n_resamples")]
fit$stability
```

`recurrence` is the inclusion frequency among repeated models.
`predictive_contribution` combines excluded-sample contribution frequency and
mutual information. `adjusted_recurrence` and `adjusted_contribution` compare
these measurements with shuffled-outcome references. The ratios define ranking
quantities. The `selected_genes` vector contains the requested top 20 genes.

```{r ranking-plot, fig.width=9, fig.height=5}
plot_gene_ranking(fit, n = 12)
```

# External evaluation

An external cohort requires the same gene identifiers and measurement scale.
All transformation parameters must come from the training cohort. A classifier
is then fitted with the selected genes in the training data and applied once to
the held-out cohort. The following outline records the required separation of
selection, fitting, and evaluation:

```{r external-evaluation, eval=FALSE}
training_fit <- select_genes(training_experiment, "severity", n_genes = 20)
selected <- training_fit$selected_genes

x_train <- t(assay(
    training_experiment[selected, ],
    "log_expression"
))
y_train <- training_experiment$severity
classifier <- glmnet::cv.glmnet(
    x_train,
    as.integer(y_train) - 1L,
    family = "binomial",
    alpha = training_fit$parameters$alpha
)
external_probability <- predict(
    classifier,
    newx = t(assay(
        external_experiment[selected, ],
        "log_expression"
    )),
    s = "lambda.min",
    type = "response"
)
pROC::roc(
    external_experiment$severity,
    as.numeric(external_probability),
    levels = levels(y_train),
    direction = "<"
)
```

This code requires preprocessed `training_experiment` and `external_experiment`
objects with aligned gene identifiers and outcome levels.

# Biological interpretation

Biological information is assessed after predictive selection. Gene Ontology
measures functional similarity, Hallmark sets identify shared biological
processes, Open Targets records disease associations, and STRING supplies
protein-association context. Database coverage and retrieval dates must
accompany interpretation. Missing annotation is recorded separately from a
measured score of zero.

The package includes biological summaries for seven transcriptomic datasets and
seven feature-selection methods. Ratios compare each selected gene set with
same-size random gene sets drawn from the same training-specific candidate
genes. A value of one is the random-set reference. The measurements are kept as
separate results because they describe different properties.

```{r biology-heatmap, fig.width=8.5, fig.height=6.3}
data(benchmark_biology)
plot_biology_comparison(
    benchmark_biology,
    method = "GeneSelectR 2.0",
    dataset_order = c(
        "Atopic dermatitis", "Bladder cancer", "Asthma", "Sepsis",
        "Psoriasis", "Tuberculosis", "Crohn's disease"
    )
)
```

For asthma, the selected-to-random ratios were 1.65 for GO similarity, 3.75 for
Hallmark sharing, 1.83 for Open Targets at a score cutoff of 0.05, and 1.56 at a
cutoff of 0.10. These values are descriptive ratios. Probability and
significance interpretations require separate analyses.

# Full-cohort asthma output

The executable fit uses 96 samples and 300 genes. The following figure uses the
full-cohort GSE69683 result extract reported in the manuscript: 323 samples, 15
outer validation divisions, and sets of 20 genes. The packaged object supports
inspection of the display without an external database request.

```{r case-study, fig.width=11, fig.height=5.5}
data(asthma_case_study)
plot_gene_evidence(asthma_case_study, comparison_label = "DGE")
```

Outer selection count records membership in each final top-20 gene set.
Within-training recurrence records inclusion among the elastic-net fits used to
construct a ranking. Adjusted contribution is compared with shuffled-outcome
fits. The Open Targets value is the recorded asthma association from release
26.06. `LINC00537` has unavailable Open Targets annotation and is marked `NA`.

# Application scope

The workflow can be transferred to other transcriptomic binary outcomes when
the following conditions are met:

1. The two outcome levels and positive class are specified before analysis.
2. Expression values have a documented, platform-appropriate normalization.
3. Candidate-gene filtering and preprocessing are estimated in training data.
4. Class sizes support the requested number of stratified folds.
5. Independent samples or outer resampling are available for performance
    estimation.
6. Biological resources match the organism and identifier system.

Targeted expression panels can also use the matrix interface when measurements
are normalized and genes have unique names. The limited candidate-gene universe
must be used for matched random-set biological comparisons.

# Reproducibility

A complete record includes the GeneSelectR 2.0 package version, input accession
and sample identifiers, phenotype definition, normalization method,
candidate-gene rule, resampling counts, random seed, alpha values, worker count,
and train-test partition. Biological analyses also require annotation-package
versions, external-resource releases, query terms, thresholds, and retrieval
dates.

The small executable result provides an interface check. The full-cohort
manuscript result has a separate provenance record and outer validation design.

# References

Friedman J, Hastie T, Tibshirani R (2010). Regularization paths for generalized
linear models via coordinate descent. *Journal of Statistical Software* 33(1),
1--22. doi:10.18637/jss.v033.i01.

Nogueira S, Sechidis K, Brown G (2018). On the stability of feature selection
algorithms. *Journal of Machine Learning Research* 18(174), 1--54.

Shaw DE et al. (2015). Clinical and inflammatory characteristics of the European
U-BIOPRED adult severe asthma cohort. *European Respiratory Journal* 46,
1308--1321. doi:10.1183/13993003.00779-2015.

# Session information

```{r session}
sessionInfo()
```
