---
title: "GeneSelectR 2.0 workflow"
author: "Damir Zhakparov"
date: "15 September 2026"
output:
    BiocStyle::html_document:
        toc: true
vignette: >
    %\VignetteIndexEntry{GeneSelectR 2.0 workflow}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

GeneSelectR 2.0 ranks genes for a binary outcome from normalized expression
measurements. Repeated elastic-net models provide gene-selection recurrence and
excluded-sample predictive contribution. Outcome permutations provide a
gene-specific reference for each quantity. The combined score defines the
ranking, and the component measurements remain available for inspection.

The package addresses predictive feature selection in transcriptomic studies.
Relevant applications include disease status, treatment response, infection
status, and pairwise molecular-subtype comparisons. Each analysis requires a
pre-specified two-level outcome. Survival, continuous-outcome, and multiclass
models require separate methods.

Differential-expression packages such as `limma`, `edgeR`, and `DESeq2` estimate
feature-wise effects and associated uncertainty. GeneSelectR 2.0 estimates a
multivariable predictive ranking. Both analyses can be reported when the study
contains inferential and predictive objectives.

# Installation

After Bioconductor acceptance, the package can be installed with:

```{r installation, eval=FALSE}
BiocManager::install("GeneSelectR2")
```

The package name used in R code is `GeneSelectR2`. Descriptive text uses
GeneSelectR 2.0.

# Study design

Model development and performance evaluation require separate sample sets.
Normalization parameters, batch-adjustment parameters, variable-gene filters,
and model tuning must be estimated from the training samples. An independent
test set or an outer resampling loop is required for a predictive-performance
estimate. The internal AUC returned by the package is used to compare elastic-
net mixing values within the training data.

The input expression values must be numeric, finite, and normalized for the
measurement platform. Count data can be transformed with an appropriate
Bioconductor workflow before gene selection. Gene identifiers must be unique
within the input object.

# Input representations

The low-level interface uses a samples-by-genes matrix and a two-level factor.
The second factor level defines the positive class.

```{r matrix-input}
library(GeneSelectR2)

set.seed(10)
n_samples <- 40
n_genes <- 20
expression_matrix <- matrix(
    rnorm(n_samples * n_genes),
    nrow = n_samples,
    ncol = n_genes,
    dimnames = list(
        paste0("sample", seq_len(n_samples)),
        paste0("gene", seq_len(n_genes))
    )
)
outcome <- factor(
    rep(c("control", "case"), each = n_samples / 2),
    levels = c("control", "case")
)
expression_matrix[outcome == "case", 1:3] <-
    expression_matrix[outcome == "case", 1:3] + 1
```

A `SummarizedExperiment` stores genes in rows, samples in columns, phenotypes in
`colData`, and provenance in `metadata`. This representation supports transfer
from other Bioconductor expression workflows.

```{r summarized-experiment}
experiment <- SummarizedExperiment::SummarizedExperiment(
    assays = list(log_expression = t(expression_matrix)),
    colData = data.frame(
        outcome = outcome,
        row.names = rownames(expression_matrix)
    )
)
dim(experiment)
SummarizedExperiment::assayNames(experiment)
```

`select_genes()` accepts this object directly. The asthma vignette provides an
evaluated `SummarizedExperiment` analysis. `geneselectr2_fit()` exposes the
matrix interface used below.

# Model fitting

The example uses reduced resampling counts to keep the vignette build short.
The `random_seed` controls resampling, outcome permutations, and model fitting.
One worker is used by default.

```{r fit}
fit <- geneselectr2_fit(
    expression_matrix,
    outcome,
    B = 5,
    permutations = 2,
    null_B = 5,
    random_seed = 123,
    workers = 1,
    verbose = FALSE
)
fit
```

The number of repeated fits, permutations, and null fits controls Monte Carlo
precision and computation time. These values should be specified in the
analysis protocol. The same values and seed should be retained for all methods
within a comparison.

# Gene-level results

`gene_scores` contains one row per input gene. `recurrence` is the fraction of
repeated models with a non-zero coefficient. `predictive_contribution` combines
excluded-sample contribution frequency and mutual information.
`adjusted_recurrence` and `adjusted_contribution` divide the observed quantities
by their gene-specific shuffled-outcome means. `final_score` defines the rank.

```{r results}
names(fit$gene_scores)
head(fit$gene_scores[, c(
    "gene", "recurrence", "predictive_contribution",
    "adjusted_recurrence", "adjusted_contribution", "final_score"
)])
fit$stability["nogueira_index"]
fit$cv_results[c("mean_auc", "sd_auc", "n_resamples")]
```

The adjusted quantities are descriptive ranking measurements. Statistical
significance and false-discovery control require a separate inferential
analysis. The Nogueira index summarizes agreement among the repeated selected
sets. Its interpretation depends on the number of candidate genes, selected-set
size, and resampling design.

```{r ranking-plot, fig.width=9, fig.height=5}
plot_gene_ranking(fit, n = 12)
```

# High-level selection

`select_genes()` compares the supplied `alpha` values by internal AUC and
returns the requested number of genes. A typical training-object call is:

```{r high-level-call, eval=FALSE}
selected_fit <- select_genes(
    training_experiment,
    outcome = "outcome",
    assay_name = "log_expression",
    n_genes = 20,
    alpha = c(0.5, 1),
    B = 50,
    permutations = 20,
    null_B = 20,
    workers = 1,
    seed = 123
)
```

An external cohort must receive preprocessing parameters estimated from the
training cohort. Prediction assessment also requires a model fitted on the
training data and a pre-specified performance metric. The gene ranking alone is
insufficient for an external-performance estimate.

# Biological assessment

Biological assessment is performed after predictive ranking. The package
supports GO semantic similarity, MSigDB membership, Open Targets disease
associations, and STRING network proximity. Annotation availability, database
versions, organism, query terms, score thresholds, and retrieval dates form
part of the analysis provenance.

GO semantic scoring can use externally specified disease terms:

```{r semantic-example, eval=FALSE}
semantic_scores <- score_semantic_layer(
    fit$gene_scores$gene,
    target_terms = c("GO:0006955", "GO:0002376"),
    ontology = "BP",
    organism = "human"
)
```

Network scoring retrieves disease-associated seed genes from Open Targets and
propagates their scores over STRING. These functions require internet access on
the first call. Retrieved resources are stored under the platform-specific R
cache directory. Cached Open Targets records include the submitted term,
resolved ontology identifier, and retrieval date.

```{r network-example, eval=FALSE}
network_scores <- score_network_layer(
    fit$gene_scores$gene,
    disease_term = "EFO_0000274",
    max_seeds = 100,
    min_score = 0.1
)
cache_info()
```

Missing annotation must remain distinguishable from measured zero evidence.
Database-derived scores are descriptive evidence measurements and require
matched random-gene-set comparisons for contextual interpretation.

# Reproducibility

The returned `parameters` element records the model settings. A complete
analysis record should also contain the package version, R version, annotation-
database versions, external-resource retrieval dates, input identifiers,
preprocessing specification, candidate-gene rule, and train-test partition.

```{r provenance}
fit$parameters
packageVersion("GeneSelectR2")
```

# References

Friedman J, Hastie T, Tibshirani R (2010). Regularization paths for generalized
linear models via coordinate descent. *Journal of Statistical Software* 33(1),
1--22. doi:10.18637/jss.v033.i01.

Nogueira S, Sechidis K, Brown G (2018). On the stability of feature selection
algorithms. *Journal of Machine Learning Research* 18(174), 1--54.

Morgan M, Obenchain V, Hester J, Pagès H (2026). *SummarizedExperiment:
SummarizedExperiment container*. Bioconductor package.

# Session information

```{r session-information}
sessionInfo()
```
