---
title: "GeneSelectR workflow"
author: "Damir Zhakparov"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc: true
vignette: >
    %\VignetteIndexEntry{GeneSelectR workflow}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

GeneSelectR ranks genes for a binary outcome. The result contains recurrence
across repeated elastic-net models, excluded-sample predictive contribution,
shuffled-outcome-adjusted measurements, and the combined ranking. Biological
interpretation is performed after gene ranking. This vignette uses a compact
simulated dataset so that the code is evaluated during package checks.

# Installation

After Bioconductor acceptance:

```{r installation, eval=FALSE}
BiocManager::install("GeneSelectR2")
```

# Input

The expression matrix has samples in rows and genes in columns. The outcome is
a two-level factor.

```{r data}
library(GeneSelectR2)

set.seed(10)
n <- 40
p <- 20
X <- matrix(rnorm(n * p), nrow = n, ncol = p)
colnames(X) <- paste0("gene", seq_len(p))
y <- factor(
    rep(c("control", "case"), each = n / 2),
    levels = c("control", "case")
)
X[y == "case", 1:3] <- X[y == "case", 1:3] + 1
```

# Model fitting

The settings below use five stratified folds. Specify the gene-set size and
resampling settings before an analysis.

```{r fit}
fit <- geneselectr2_fit(
    X,
    y,
    B = 5,
    permutations = 2,
    null_B = 5,
    verbose = FALSE
)
```

# Results

`gene_scores` contains one row per input gene. The table retains every
measurement used to create the ranking.

```{r results}
head(fit$gene_scores[, c(
    "gene", "recurrence", "adjusted_contribution", "final_score"
)])
fit$stability["nogueira_index"]
fit$cv_results[c("mean_auc", "sd_auc", "n_resamples")]
```

# Reproducibility

The returned `parameters` element records the model settings. Analyses should
also record the package version, R version, annotation-database versions, and
external-resource retrieval dates.

```{r provenance}
fit$parameters
sessionInfo()
```
